---
title: "Theoretical modelling the early phase of the Belgian COVID-19 epidemic using
  compartmental model and studying the factors associated with the hospitalization
  dynamics"
author: "Mariam ISMAIL"
date: "2025-04-09"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The COVID-19 pandemic led to the generation of an unprecedented volume of global data. This project focuses on i) developing a theoretical model using an extended SIR model and ii) studying factors associated with the number of new hospitalization patients during the early phase of the Belgian COVID-19 epidemic.

## Data

The dataset used in this study is the COVID-19 Open Dataset (COD), available at goo.gle/covid-19-open-data. This dataset provides comprehensive time series data from multiple sources and includes two main categories of information. First, it contains epidemiological data such as the number of new hospitalized patients, confirmed cases, deaths, and recoveries. Second, it includes time series data on potentially relevant predictors of new hospitalizations, such as population mobility patterns and weather conditions. This combination allows for an analysis of the factors associated with the hospitalization dynamics. For this study, data from Belgium was selected, focusing on the early phase of the epidemic between March 15, 2020, and June 29, 2020.

## Analysis

### Import necessary libraries and the data

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(mgcv)   
library(lubridate)
library(tidyr)
library(deSolve)
library(reshape2)

# Define filenames and corresponding object names
files <- c(
  "epidemiology.csv",
  "weather.csv",
  "hospitalizations.csv",
  "mobility.csv"
)

names(files) <- c(
  "epidemiology",
  "weather",
  "hospitalizations",
  "mobility"
)

# Define the base path
base_path <- "data/belgium_clean"

# Loop through the files
for (name in names(files)) {
  path <- file.path(base_path, files[name])
  
  # Read the CSV file
  df <- read.csv(path, sep = ";")
  
  # Filter for Belgium 
  if ("location_key" %in% names(df)) {
    df <- df %>% filter(location_key == "BE")
  }
  
  # Convert date column to Date format and sort 
  if ("date" %in% names(df)) {
    df$date <- as.Date(df$date, format = "%d/%m/%Y")
    df <- df %>% arrange(date)
  }
  
  # Assign cleaned data to variable in environment
  assign(name, df)
}
```

### Merge the data into one dataframe

```{r echo=TRUE, message=FALSE, warning=FALSE}
df1 <- merge(epidemiology, weather, by = c("date", "location_key"))
df2 <- merge(df1, hospitalizations, by = c("date", "location_key"))
Belgium_data <- merge(df2, mobility, by = c("date", "location_key"))
Belgium_data_initial_phase <- Belgium_data %>% filter(date >= as.Date("2020-03-15") & date <= as.Date("2020-06-29")) # selecting the preiod for the initial phase
summary(Belgium_data_initial_phase$date)
```

### Part 1: Developing a theoretical model using an extended SIR model

```{r seih-diagram, echo=FALSE}
library(DiagrammeR)

grViz("
digraph SEIH {
  graph [layout = dot, rankdir = LR]

  node [shape = circle, style=filled, fillcolor=lightblue]
  S [label = 'S']
  E [label = 'E']
  I [label = 'I']
  R [label = 'R']
  H [label = 'H']
  D [label = 'D']

  S -> E [label = 'λ']
  E -> I [label = 'ε']
  I -> H [label = 'σ']
  I -> R [label = 'μ']
  H -> R [label = 'τ']
  H -> D [label = 'θ']

}
")
```
**Fig. 1** Schematic representation of the extended SIR compartmental model. Individuals begin in the susceptible compartment (S) and become exposed (E) through contact with infected individuals (I). After a latent period, they transition to the infected state. Infected individuals with severe symptoms may be hospitalized (H). The possible outcomes from infection or hospitalization are either recovery (R) or death (D).


```{r echo=TRUE, message=FALSE, warning=FALSE}
# Initial number of people in each compartment
initial_state_values <- c(S = 11455519 - 25000*5 - 14480,  # susceptible
                          E = 25000*5,                     # exposed
                          I = 14480,                       # infected
                          R = 0,                           # recovered
                          H = 0,                           # hospitalized
                          D = 0)                           # deceased


# Parameters describing the transition rates in units of days^-1
parameters <- c(beta = 0.0999,       # transmission probability
                epsilon = 0.714,     # rate from exposed to infected
                sigma = 0.25,        # rate from infected to hospitalized 
                mu = (1/2.4)*(1/2.4), # rate from infected to recovered
                tau = 1/5,           # rate from hospitalized to recovered
                theta = 1/4 )        # rate from hospitalized to death
            
# Sequence of timesteps 
times <- seq(from = 0, to = 121, by = 1) # from 0 to 121 days in daily intervals

# Extended SIR Model
compartmental_model <- function(time, state, parameters) {  

    with(as.list(c(state, parameters)), {    
        
    # Calculating the total population size N 
      N <- S+E+I+R+H+D    # total population            
      
    # defining lambda
      lambda <- beta * I/N    # force of infection
        
    # The differential equations
      dS <- -lambda * S
      dE <- lambda * S - epsilon*E
      dI <- epsilon * E - mu * I - sigma * I    
      dH <- sigma * I - tau * H - theta * H
      dR <- mu * I + tau * H          
      dD <- theta*H                                    
    return(list(c(dS, dE, dI, dH, dR, dD)))         
    })
  
}

# Solving the ode
compartmental_output <- as.data.frame(ode(y = initial_state_values, 
                            times = times, 
                            func = compartmental_model,
                            parms = parameters))

# Plot the predicted number of new hospitalized patients
ggplot(data = compartmental_output,                                               
       aes(x = time, y = parameters["sigma"]*compartmental_output$I)) +     
                                     
  geom_line() +    
  xlab("Time (days)")+                                                   
  ylab("New hospitalized patients") +                                      
  labs(title = "Model new hospitalized patients per day")
```

### Plot the observed new hospitalized patients 

```{r echo=TRUE, message=FALSE, warning=FALSE}
Belgium_data_long <- pivot_longer(Belgium_data_initial_phase, cols = c(new_hospitalized_patients), names_to = "variable", values_to = "count") 

ggplot(Belgium_data_long, aes(x = Belgium_data_long$date, y = count, color = variable)) +  
geom_point(method = "loess", se = FALSE) + labs(title = "Observed new hospitalized patients", x = "Date", y = "Count") + theme_minimal()
```

### Plot the model and observed new hospitalized patients

```{r}
model_df <- data.frame(
  time = compartmental_output$time,
  count = parameters["sigma"] * compartmental_output$I,
  source = "Model"
)
Belgium_data_long <- pivot_longer(
  Belgium_data_initial_phase,
  cols = c(new_hospitalized_patients),
  names_to = "variable",
  values_to = "count"
)

observed_df <- data.frame(
  time = as.numeric(as.Date(Belgium_data_long$date) - min(as.Date(Belgium_data_long$date))),  # convert date to numeric days
  count = Belgium_data_long$count,
  source = "Observed"
)

combined_df <- rbind(model_df, observed_df)
ggplot(combined_df, aes(x = time, y = count, color = source)) +
  geom_line(data = subset(combined_df, source == "Model"), size = 1) +
  geom_point(data = subset(combined_df, source == "Observed"), alpha = 0.7) +
  xlab("Time (days)") +
  ylab("New hospitalized patients") +
  labs(title = "Modeled vs observed new hospitalized patients") +
  theme_minimal()
```

### Part 2: studying factors associated with the number of new hospitalization patients  

The objective of this part is to identify variables associated with the number of new hospitalization patients to guide future caliberation steps with more realistic assumptions. A Poisson regression model is used, with the number of new hospitalized patients as the primary outcome variable in the analysis.

### Model 2.1: effect of temperature on hospitalizations

```{r}
# Poisson model
model1 <- glm(new_hospitalized_patients ~ average_temperature_celsius ,
              family = poisson(link = "log"),
data = Belgium_data_initial_phase)
summary(model1)
Belgium_data_initial_phase$fitted1 <- predict(model1, type = "response")

# Plot
ggplot(Belgium_data_initial_phase, aes(x = date)) +
  geom_line(aes(y = new_hospitalized_patients, color = "Observed")) +
  geom_line(aes(y = fitted1, color = "Predicted")) +
  scale_color_manual(values = c("Observed" = "blue", "Predicted" = "red")) +
  labs(
    title = "Modeling hospitalizations as a function of temperature: observed vs predicted values",
    y = "Hospitalizations",
    color = "Legend" ) +
  theme_minimal()

```

This model captures the general shape but underperforms in several areas (notably underestimating peaks and overestimating in low hospitalization periods). 

### Model 2.2: effect of lagged temperature on hospitalizations

```{r}
# add lag dates
Belgium_data_initial_phase$temp_lag7 <- dplyr::lag(Belgium_data_initial_phase$average_temperature_celsius, 7) 

# remove the 7 empty values created
Belgium_data_initial_phase_clean <- Belgium_data_initial_phase %>% filter(!is.na(temp_lag7)) 

# Poisson model
model2 <- glm(new_hospitalized_patients ~ temp_lag7,
              family = poisson(link = "log"),
              data = Belgium_data_initial_phase_clean)
summary(model2)

Belgium_data_initial_phase_clean$fitted2 <- predict(model2, type = "response")

# Plot
ggplot(Belgium_data_initial_phase_clean, aes(x = date)) +
  geom_line(aes(y = new_hospitalized_patients, color = "Observed")) +
  geom_line(aes(y = fitted2, color = "Predicted")) +
  scale_color_manual(values = c("Observed" = "blue", "Predicted" = "red")) +
  labs(
    title = "Modeling hospitalizations as a function of lagged temperature: observed vs predicted values",
    y = "Hospitalizations",
    color = "Legend" ) +
  theme_minimal()

```

This model has slight improvement over non-lagged temperature, particularly in peak alignment.

### Model 2.3: effect of relative humidity on hospitalizations

```{r}
# Poisson model
model3 <- glm(new_hospitalized_patients ~ relative_humidity,
              family = poisson(link = "log"),
              data = Belgium_data_initial_phase)
summary(model3)
Belgium_data_initial_phase$fitted3 <- predict(model3, type = "response")

# Plot
ggplot(Belgium_data_initial_phase, aes(x = date)) +
  geom_line(aes(y = new_hospitalized_patients, color = "Observed")) +
  geom_line(aes(y = fitted3, color = "Predicted")) +
  scale_color_manual(values = c("Observed" = "blue", "Predicted" = "red")) +
  labs(
    title = "Modeling hospitalizations as a function of relative humidity: Observed vs Predicted Values",
    y = "Hospitalizations",
    color = "Legend"
  ) +
  theme_minimal()

```

This model suggests that humidity alone has weak explanatory power.

### Model 2.4: effect of temperature on hospitalizations with relative humadity as a covariate

```{r}
# Poisson model
model4 <- glm(new_hospitalized_patients ~ average_temperature_celsius + relative_humidity,
              family = poisson(link = "log"),
              data = Belgium_data_initial_phase)
summary(model4)
Belgium_data_initial_phase$fitted4 <- predict(model4, type = "response")

# Plot
ggplot(Belgium_data_initial_phase, aes(x = date)) +
  geom_line(aes(y = new_hospitalized_patients, color = "Observed")) +
  geom_line(aes(y = fitted4, color = "Predicted")) +
  scale_color_manual(values = c("Observed" = "blue", "Predicted" = "red")) +
  labs(
    title = "Modeling hospitalizations as a function of temperature with relative humidity as a covariate: Observed vs Predicted values",
       y = "Hospitalizations",
    color = "Legend"
  ) +
  theme_minimal()

```

This model shows improved alignment compared to using temperature or humidity alone.

### Model 2.5: effect of mobility on hospitalizations 


```{r}
# Poisson model
model5 <- glm(new_hospitalized_patients ~ mobility_retail_and_recreation + mobility_grocery_and_pharmacy + mobility_parks + mobility_transit_stations + mobility_workplaces + mobility_residential , family = poisson(link = "log"), data = Belgium_data_initial_phase)
summary(model5)
Belgium_data_initial_phase$fitted5 <- predict(model5, type = "response")

# Plot
ggplot(Belgium_data_initial_phase, aes(x = date)) +
  geom_line(aes(y = new_hospitalized_patients, color = "Observed")) +
  geom_line(aes(y = fitted5, color = "Predicted")) +
  scale_color_manual(values = c("Observed" = "blue", "Predicted" = "red")) +
  labs(
    title = "Modeling hospitalizations as a function of mobility variables: observed vs predicted values",
       y = "Hospitalizations",
    color = "Legend"
  ) +
  theme_minimal()
```

This model shows strong fit, especially during the peak and decay periods.

### Model 2.6: effect of mobility on hospitalizations - PCA approach

### PCA on COVID-19 mobility variables

```{r}
mobility_vars <- Belgium_data_initial_phase[, c("mobility_retail_and_recreation", "mobility_grocery_and_pharmacy",
                          "mobility_parks", "mobility_transit_stations",
                          "mobility_workplaces", "mobility_residential")]

mobility_scaled <- scale(mobility_vars)

pca_result <- prcomp(mobility_scaled, center = TRUE, scale. = TRUE)

summary(pca_result)
screeplot(pca_result, type = "lines")
title(xlab = "Principal Components")
```

```{r}
Belgium_data_initial_phase$PC1 <- pca_result$x[, 1]
Belgium_data_initial_phase$PC2 <- pca_result$x[, 2]
model6 <- glm(new_hospitalized_patients ~ PC1 + PC2 , family = poisson(link = "log"), data = Belgium_data_initial_phase)
summary(model6)

# Poisson model
Belgium_data_initial_phase$fitted6 <- predict(model6, type = "response")

# Plot
ggplot(Belgium_data_initial_phase, aes(x = date)) +
  geom_line(aes(y = new_hospitalized_patients, color = "Observed")) +
  geom_line(aes(y = fitted6, color = "Predicted")) +
  scale_color_manual(values = c("Observed" = "blue", "Predicted" = "red")) +
  labs(
    title = "Modeling hospitalizations as a function of mobility indicators: observed vs predicted values",
       y = "Hospitalizations",
    color = "Legend"
  ) +
  theme_minimal()

```

This model shows some instability or outliers in predicted values, possibly due to model overfitting or noisy data.

### Model 2.7: effect of temperature on hospitalizations with mobility variables as covariates

```{r}
# Poisson model
model7 <- glm(new_hospitalized_patients ~ average_temperature_celsius + mobility_retail_and_recreation + mobility_grocery_and_pharmacy + mobility_parks + mobility_transit_stations + mobility_workplaces + mobility_residential, family = poisson(link = "log"), data = Belgium_data_initial_phase)
summary(model6)
Belgium_data_initial_phase$fitted7 <- predict(model7, type = "response")

# Plot
ggplot(Belgium_data_initial_phase, aes(x = date)) +
  geom_line(aes(y = new_hospitalized_patients
), color = "blue") +
  geom_line(aes(y = fitted7), color = "red") +
  labs(title = "Modeling hospitalizations as a function of mobility indicators with temperature as a covariate: observed vs predicted values",
       y = "Hospitalizations")
```

This model captures both peak dynamics and low periods reasonably well. It demonstrates that combining mobility  with environmental control variables (temperature) gives the best predictive power.

## Conclusion

The comparison of observed versus predicted hospitalizations across different covariate models demonstrates that mobility data is the most reliable predictor of hospitalization trends, particularly when combined with temperature as a control variable. Models based on temperature or humidity alone are insufficient to capture the complexity of epidemic dynamics. However, the inclusion of lagged environmental effects and behavioral indicators greatly improves prediction accuracy. These insights suggest that an extended SIR model calibrated with dynamic, multi-factor inputs—especially mobility and weather—is essential for accurately modeling hospitalization trends. Incorporating these variables allows for a more realistic and adaptable model that can inform public health decisions during evolving epidemic scenarios.



